---
title: "Tutorial 5. CCC Analysis with scRNA-seq dataset"
author: "Iguaracy Souza and Umran YAMAN"
date: 'Compiled: `r Sys.Date()`'  
output: 
  html_document:
    toc: TRUE
    code_folding:  "show"
    theme: united
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  warning = FALSE,
  error = TRUE,
  echo = TRUE,
  message = FALSE,
  fig.width = 8,
  time_it = TRUE
)
```

### Data and library import

Load necessary libraries, and read in the preprocessed scRNA-seq dataset.

```{r}
suppressPackageStartupMessages({
  library("Seurat")
  library("tidyverse")
  library("magrittr")
  library("liana")
  library("dplyr")
  library("ggplot2")
  library("patchwork")
  library("knitr")
})
```

```{r}
scRNA_brain <- readRDS("../data_to_delete/scRNA_brain_seurat_QC.rds")
```

### Split the dataset

Split the data by 'Pathology' to separate cells based on their pathology status, since we would like to see CCC in through the disease progression.

```{r}
Idents(scRNA_brain) <- scRNA_brain$major.celltype
data_split <- SplitObject(scRNA_brain, split.by = "Pathology")
```

### Run LIANA 

Initialize a list to store LIANA results for each subset.


```{r}
liana_results <- list()

# Run LIANA for each subset
for (subset_name in names(data_split)) {
  print(paste("Processing", subset_name))
  
  # Assuming that the LIANA wrap function works directly on a Seurat object
  liana_subset_result <- liana_wrap(data_split[[subset_name]])
  
  # Store the result
  liana_results[[subset_name]] <- liana_subset_result
  
}
```

### Aggregate the LIANA results 

Now, liana_results contains all the LIANA results for each cell type. Initialize an empty list for storing aggregated results for all pathologies.


```{r}
aggregated_liana_all <- list()
# Process and save results for each subset
for (subset_name in names(liana_results)) {
  print(paste("Aggregating and filtering results for", subset_name))
  
  # Assuming liana_aggregate() and the filtering are correctly applied on your liana results
  # Aggregate results
  aggregated_liana <- liana_results[[subset_name]] %>% 
    liana_aggregate()
  
  # Filter results
  filtered_liana <- aggregated_liana %>%
    dplyr::filter(aggregate_rank <= 0.05)  # Ensure dplyr is specified to avoid method conflicts
  
  # Store the aggregated (and optionally filtered) results in the list
  aggregated_liana_all[[subset_name]] <- filtered_liana
  
  # Save the filtered results to a CSV file
  csv_filename <- paste0("../data_to_delete/Cell_cell_results/LIANA_scRNA_Brain_", subset_name, ".csv")
  write.csv(filtered_liana, csv_filename, row.names = FALSE)
  
  print(paste("Results saved for", subset_name, "in", csv_filename))
}
```

### Check cell-cell interactions by pathology

In healthy cases, we have the microglia communicating microglia and vasculature only, probably due to protect homoestatis. Signatures are given here is not AD pathology related.

```{r}
major.celltype <- levels(scRNA_brain$major.celltype)
major.celltype

# Specify the cell type of interest
cell_type_of_interest <- "Mic"  # Replace with the cell type(s) of you're interest

##cell sending the signal (Ligand)
###Non AD
# Adjust the function call based on how 'liana_dotplot' is implemented and expects its arguments - sender
dotplot_Mic_source_nonAD <- liana_dotplot(
  aggregated_liana_all$nonAD,
  source_groups = c(cell_type_of_interest),  # Adjust this based on how your cell types are categorized (source or target)
  target_groups = NULL,  # Adjust target groups as needed or based on your specific analysis requirements
  ntop = 20  # Assuming you want the top 20 interactions
)

tiff("../figures/dotplot_Mic_source_nonAD.tiff", units="in", width=10, height=15, res=300)
dotplot_Mic_source_nonAD
invisible(dev.off())
include_graphics("../figures/dotplot_Mic_source_nonAD.tiff")
```

We move onto the mild AD pathology, where mainly amyloid pathology burden is seen in lower degrees. We can say that this microglial response is due to the amyloid response in early AD. We now see that the cell types microglia interact is increasing.  


```{r}
##Early AD
dotplot_Mic_source_earlyAD <- liana_dotplot(
  aggregated_liana_all$earlyAD,
  source_groups = c(cell_type_of_interest),  # Adjust this based on how your cell types are categorized (source or target)
  target_groups = NULL,  # Adjust target groups as needed or based on your specific analysis requirements
  ntop = 20  # Assuming you want the top 20 interactions
)
tiff("../figures/dotplot_Mic_source_earlyAD.tiff", units="in", width=10, height=15, res=300)
dotplot_Mic_source_earlyAD
invisible(dev.off())
include_graphics("../figures/dotplot_Mic_source_earlyAD.tiff")
```


```{r}
##Late AD
dotplot_Mic_source_lateAD <- liana_dotplot(
  aggregated_liana_all$lateAD,
  source_groups = c(cell_type_of_interest),  # Adjust this based on how your cell types are categorized (source or target)
  target_groups = NULL,  # Adjust target groups as needed or based on your specific analysis requirements
  ntop = 20  # Assuming you want the top 20 interactions
)

tiff("../figures/dotplot_Mic_source_lateAD.tiff", units="in", width=10, height=15, res=300)
dotplot_Mic_source_lateAD
invisible(dev.off())
include_graphics("../figures/dotplot_Mic_source_lateAD.tiff")

tiff("../figures/dotplot_Mic_source_allAD.tiff", units="in", width=12, height=30, res=300)
(dotplot_Mic_source_nonAD | dotplot_Mic_source_earlyAD | dotplot_Mic_source_lateAD) +
  plot_annotation(title = "LIANA Dot Plots: Non AD vs Early AD vs Late AD")
invisible(dev.off())
include_graphics("../figures/dotplot_Mic_source_allAD.tiff")
```


### Adjust Liana plots - Try it yourself!

Adjust the function call based on how 'liana_dotplot' is implemented and expects its arguments -between  sender (ligand) and receiver (receptor)

```{r}
###Non AD
dotplot_Mic_target_nonAD <- liana_dotplot(
  aggregated_liana_all$nonAD,
  source_groups = NULL,  # Adjust this based on how your cell types are categorized (source or target)
  target_groups = c(cell_type_of_interest),  # Adjust target groups as needed or based on your specific analysis requirements
  ntop = 20  # Assuming you want the top 20 interactions
)

tiff("../figures/dotplot_Mic_source_nonAD_adjust.tiff", units="in", width=10, height=15, res=300)
dotplot_Mic_target_nonAD
invisible(dev.off())
include_graphics("../figures/dotplot_Mic_nonAD_adjust.tiff")

##Early AD
dotplot_Mic_target_earlyAD <- liana_dotplot(
  aggregated_liana_all$earlyAD,
  source_groups = NULL,  # Adjust this based on how your cell types are categorized (source or target)
  target_groups = c(cell_type_of_interest),  # Adjust target groups as needed or based on your specific analysis requirements
  ntop = 20  # Assuming you want the top 20 interactions
)

tiff("../figures/dotplot_Mic_source_earlyAD_adjust.tiff", units="in", width=10, height=15, res=300)
dotplot_Mic_target_earlyAD
invisible(dev.off())
include_graphics("../figures/dotplot_Mic_earlyAD_adjust.tiff")

##Late AD
dotplot_Mic_target_lateAD <- liana_dotplot(
  aggregated_liana_all$lateAD,
  source_groups = NULL,  # Adjust this based on how your cell types are categorized (source or target)
  target_groups = c(cell_type_of_interest),  # Adjust target groups as needed or based on your specific analysis requirements
  ntop = 20  # Assuming you want the top 20 interactions
)

tiff("../figures/dotplot_Mic_source_lateAD_adjust.tiff", units="in", width=10, height=15, res=300)
dotplot_Mic_target_lateAD
invisible(dev.off())
include_graphics("../figures/dotplot_Mic_lateAD_adjust.tiff")

tiff("../figures/dotplot_Mic_source_allAD_adjust.tiff", units="in", width=10, height=15, res=300)
(dotplot_Mic_target_nonAD | dotplot_Mic_target_earlyAD | dotplot_Mic_target_lateAD) +
  plot_annotation(title = "LIANA Dot Plots: Non AD vs Early AD vs Late AD")
invisible(dev.off())
include_graphics("../figures/dotplot_Mic_allAD_adjust.tiff")
```



